Refactoring the \texttt{nls()} function in R

### 2023-01-09 RefactoringNLS/RefactoringNLS.tex
reedits Jan 2023

\texttt{nls()} is the tool in base R, the primary software software
distribution from the Comprehensive R Archive Network
(\url{https://cran.r-project.org}) for estimating nonlinear statistical
models. The function dates to the 1980s and the work related to Bates
and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}).

@@

\begin{verbatim}
## residual sumsquares =  0  on  11 observations
##     after  1    Jacobian and  1 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## a                     -3            NA         NA         NA           0       26.49  
## b                      5            NA         NA         NA           0       9.615  
## c                    0.5            NA         NA         NA           0       2.466  
## r                    0.7            NA         NA         NA           0      0.1098  
## m                      1            NA         NA         NA           0   9.798e-17
\end{verbatim}

@@

\texttt{nls()}, with the \texttt{numericDeriv()} function, computes the
Jacobian as the ``gradient'' attribute of the residual vector. This is
implemented as a mix of R and C code, but we have created a rather more
compact version entirely in R. See
\url{https://github.com/nashjc/RNonlinearLS/DerivsNLS} .

As far as we can understand the logic in \texttt{nls()}, the Jacobian
computation during parameter estimation is carried out entirely within
the called C-language program, and the R code function
\texttt{numericDeriv()}, part of \texttt{./src/library/stats/R/nls.R} in
the R distribution source code. This is used to provide Jacobian
information in the \texttt{nlsModel()} and \texttt{nlsModel.plinear()}
functions, which are not exported for general use.

@@

Note that the selfStart models in the base R
\texttt{./src/library/stats/R/zzModels.R} file may provide the Jacobian
in the ``gradient'' attribute of the ``one-sided'' formula that defines
each model, and these Jacobians may be the analytic forms. The
\texttt{nls()} function, after computing the ``right hand side'' or
\texttt{rhs} of the residual, then checks to see if the ``gradient''
attribute is defined, and, if not, uses \texttt{numericDeriv} to compute
a Jacobian into that attribute. This code is within the
\texttt{nlsModel()} or \texttt{nlsModel.plinear()} functions. The use of
analytic Jacobians almost certainly contributes to the good performance
of \texttt{nls()} on selfStart models.

@@

?? does resid return weighted or unweighted residuals

?? do we want unweighted (raw) residuals? or are predictions / fits just
as good?

@@

While the implementation of subset at the level of the call to
\texttt{nls()} has a certain attractiveness, it does mean that the
programmer of the solver needs to be aware of the source (and value) of
objects such as the data, residuals and Jacobian. By preference, we
would implement subsetting by means of zero-value weights, with
observation counts (and degrees of freedom) computed via the numbers of
non-zero weights. Alternatively, we would extract a working dataframe
from the relevant elements in the original.

@@

\texttt{na.action} is an argument to the \texttt{nls()} function, but it
does not appear obviously in the source code, often being handled behind
the scenes after referencing the option \texttt{na.action}. A useful,
but possibly dated, description is given in:
\url{https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/}.

@@

A succinct description of the issue is given in:
\emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}

@@

\texttt{nls()} can be called without specifying the \texttt{data}
argument. In this case, it will search in the available environments
(i.e., workspaces) for suitable data objects. We do NOT like this
approach, but it is ``the R way''. R allows users to leave many objects
in the default (.GlobalEnv) workspace. Moreover, users have to actively
suppress saving this workspace (\texttt{.RData}) on exit; otherwise, any
such file in the path, when R is launched, will be loaded. The
overwhelming proportion of R users in our acquaintance avoid saving the
workspace because of the danger of lurking data and functions which may
cause unwanted results.

@@

Nonlinear estimation algorithms are almost all iterative and need a set
of starting parameters. \texttt{nls()} offers a special class of
modeling functions called \textbf{selfStart} models. There are a number
of these in base R (\texttt{./src/library/stats/R/zzModels.R}) and
others in R packages such as CRAN package \texttt{nlraa} (Miguez
(\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
now-archived package \texttt{NRAIA}. Unfortunately, some
\textbf{selfStart} codes entangle the calculation of starting values for
parameters with the particulars of the \texttt{nls()} code. Though there
is a \texttt{getInitial()} function, this is not easy to use to simply
compute the initial parameter estimates outside of \texttt{nls()}, in
part because it may call that function. Such circular references are, in
our view, dangerous. Moreover, we believe that it would be helpful to
have selfStart models that allow users to explicitly provide the
starting parameters.

@@

Because the Gauss-Newton algorithm is unreliable from many sets of
starting of parameters, selfStart models are a part of the
\texttt{nls()} infrastructure rather than an accessory. However,
creating such functions is a lot of work, and their documentation (file
\texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated. We
believe that the focus should be placed on getting good initial
parameters, that is \texttt{getInitial()} function, though avoiding the
current calls back to \texttt{nls()}. Interactive tools, such as
``visual fitting'' (Nash and Velleman
(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
considering.

@@

The result object from \texttt{nlsr::nlxb()} is still much smaller than
the one \texttt{nls()} returns. Moreover, \texttt{nlxb} explicitly
returns the sum of squares as well as the residual vector and Jacobian.
The counts of evaluations are also returned. (Note that the singular
values of the Jacobian are actually computed in the \texttt{print} and
\texttt{summary} methods for the result.) We noted several potential
updates to the \texttt{nlsr} documentation as well as that for
\texttt{nls()} as we made our comparisons.

@@

The function \texttt{resid()} (an alias for \texttt{residuals()})
\texttt{nlsModel()} is WEIGHTED, as are those in the structured object
\texttt{m} returned by \texttt{nls()} or \texttt{minpack.lm::nlsLM},
e.g., \texttt{ans\$m\$resid()}.

@@

As the \texttt{nls()} \textbf{man} page states, when the ``port''
algorithm is used with the \texttt{trace} argument TRUE, the iterations
display the objective function value which is 1/2 the sum of squares (or
deviance). It is likely that the trace display is embedded in the
Fortran of the \texttt{nlminb} routine that is called to execute the
``port'' algorithm, but the discrepancy is nonetheless unfortunate for
users.

@@

Moreover, specifying a model to a solver should, ideally, use the same
syntax across solver tools. Unfortunately, R allows multiple approaches
within different modelling tools, and within \texttt{nls()} itself. This
is highlighted by the case of partially linear models.

@@

\begin{verbatim}
## > with(mm, table(Strip))
## Strip
## S01 S02 S03 
##   4   4   4 
## > nl <- nlevels(mm$Strip)
## > musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, start = list(th = 1), 
## +     algorithm = "plinear")
## > summary(musc.1)
## 
## Formula: Length ~ cbind(1, exp(-Conc/th))
## 
## Parameters:
##       Estimate Std. Error t value Pr(>|t|)    
## th       0.624      0.222    2.81   0.0203 *  
## .lin1   25.684      1.441   17.82  2.5e-08 ***
## .lin2  -26.631      6.147   -4.33   0.0019 ** 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 2.96 on 9 degrees of freedom
## 
## Number of iterations to convergence: 5 
## Achieved convergence tolerance: 4.71e-07
## 
## > b <- coef(musc.1)
## > musc.2 <- nls(Length ~ a[Strip] + b[Strip] * exp(-Conc/th), data = mm, 
## +     start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
## > summary(musc.2)
## 
## Formula: Length ~ a[Strip] + b[Strip] * exp(-Conc/th)
## 
## Parameters:
##    Estimate Std. Error t value Pr(>|t|)    
## a1  22.9277     0.5154   44.49  1.1e-07 ***
## a2  27.8606     0.5150   54.09  4.1e-08 ***
## a3  28.3426     1.0771   26.31  1.5e-06 ***
## b1 -44.1558    12.3900   -3.56    0.016 *  
## b2 -43.8084    12.3148   -3.56    0.016 *  
## b3 -32.8421     2.1553  -15.24  2.2e-05 ***
## th   0.5548     0.0808    6.86    0.001 ** 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.798 on 5 degrees of freedom
## 
## Number of iterations to convergence: 6 
## Achieved convergence tolerance: 2.03e-06
\end{verbatim}

@@

We want:

@@

These goals echo themes in the recent discussion Vidoni
(\protect\hyperlink{ref-Vidoni21}{2021}) and accompanying commentary.

\hypertarget{rationalization-of-formula-specifications}{%
\subsubsection{Rationalization of formula
specifications}\label{rationalization-of-formula-specifications}}

We have seen that \texttt{nls()} uses a different formula specification
from the default if the \texttt{plinear} algorithm is used. This is
unfortunate, since the user cannot then simply add
\texttt{algorithm="plinear"} to the call, making errors more likely.

Nonlinear models, apart from those using the ``plinear'' algorithm, need
all parameters specified. That is, the model formula should be written
down as it would be computed. Thus a linear model (which is valid input
for nonlinear estimation) should be specified
\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} rather than
\texttt{y\ \textasciitilde{}\ x}. Moreover, we want to avoid the
partially linear parameters appearing in the result object as
\texttt{.lin1}, \texttt{.lin2}, etc., where the actual position of
parameters in the model is not explicit. We are unclear why these
parameters begin with a dot, that is, that they have the ``hidden''
indicator.

@@

\hypertarget{rationalization-of-indexed-models}{%
\subsubsection{Rationalization of indexed
models}\label{rationalization-of-indexed-models}}

@@

\hypertarget{streamlining-code}{%
\subsubsection{Streamlining code}\label{streamlining-code}}

@@

Such goals align with the aims of \textbf{unit testing} (e.g.,
\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211},
Wickham (\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
(\protect\hyperlink{ref-HWdevtools21}{2021}) and the conventional R
package testing tools).

\hypertarget{a-testing-package-nlscompare}{%
\subsubsection{A testing package:
NLSCompare}\label{a-testing-package-nlscompare}}

One of us has developed a working prototype package at
\url{https://github.com/ArkaB-DS/nlsCompare} . A primary design
objective of this is to allow the summarisation of multiple tests in a
compact output. The prototype has a vignette to illustrate its use.

@@

In our investigation, we built several resources, which are now part of
the repository \url{https://github.com/nashjc/RNonlinearLS/}. Particular
items are:

\begin{itemize}
\item
  A BibTex bibliography for use with all documents in this project, but
  which has wider application to nonlinear least squares projects in
  general
  (\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib})
\item
  \texttt{MachID.R} offers a suggested concise summary function to
  identify a particular computational system used for tests. A
  discussion of how it was built and the resources needed across
  platforms is given in at
  \url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}
\item
  Nash and Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022})
  is an explanation of the construction of the \texttt{nlspkg} from the
  \texttt{nls()} code in R-base.
\item
  As the 2021 Summer of Code period was ending, one of us (JN) was
  invited to prepare a review of optimization in R. Ideas from the
  present work have been instrumental in the creation of Nash
  (\protect\hyperlink{ref-NashWires22}{2022}).
\end{itemize}

@@

It is possible that \texttt{nls()} will remain more or less as it has
been for the past two decades. Given its importance to R, a focus of
discussion should be what needs fixing and how that should be
approached, and we welcome an opportunity to participate in such
conversations.

From a design point of view, a key difference in approach between
\texttt{nls()} and \texttt{nlsr::nlxb()} is that \texttt{nls()} builds a
large infrastructure, especially the \texttt{nlsModel()} function, from
which the Gauss-Newton iteration can be executed and other statistical
information such as profiles can be computed, while \texttt{nlxb()}
returns quite limited information, and computes what is needed on an
``as and when'' basis. This follows a path that one of us (JN)
established almost 50 years ago with the software that became Nash
(\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
post-solution analysis phases of computations.

\hypertarget{opinions}{%
\subsection{Opinions}\label{opinions}}

To advance the stability and maintainability of R, we believe the
program objects (R functions) that are created by tools such as
\texttt{nls()} should have minimal interactions and side-effects. The
aspects of \texttt{nls()} that have given us the most trouble:

\begin{itemize}
\item
  The functions that compute the residuals and jacobians often presume
  the data and parameters needed are available in a particular
  environment. As long as the correct environment is used, this provides
  a compact syntax to invoke the calculations. The danger is that the
  wrong data is accessed if the internal search finds a valid name that
  is not the object we want.
\item
  Weights, subsets, and various contextual controls (such as that for
  the \texttt{na.action} option) are similarly taken from the first
  available source. Again, this makes for a very simple invocation of
  calculations, but the context is hidden from the user. Furthermore, it
  can be difficult for those of us trying to maintain or improve the
  code to be certain we have the context correct.
\item
  The mixing of R, C and Fortran code was necessary for computational
  performance in the past. Lacking decent developer documentation,
  programmers now face a lot of work to fix or improve code. We believe
  in having at least a working reference version of code in a single
  programming language. If necessary, by measuring (``profiling'') the
  code, we can find bottlenecks and substitute for those slower parts of
  the reference code.
\item
  A design that isolates the setup, solution, and post-solution parts of
  for complicated calculations reduces the number of objects that must
  be kept in alignment. On the other hand, it may mean users have to add
  more commands to complete their computations.
\item
  All iterative codes should return the best solution they have found so
  far, even if there are untoward conditions reached, such as a singular
  Jacobian. This modification could be made to \texttt{nls()} in the
  short term, correcting the issue identified above under ``Failure to
  return best result achieved''.
\item
  Given the existence of examples of good practices such as analytic
  derivatives, stabilized solution of Gauss-Newton equations and
  bounds-constrained parameters, base R tools should be moving to
  incorporate them.
\end{itemize}

@@

Hans Werner Borchers has been helpful in developing the proposal for
this project and in comments on this and related work. Heather Turner
co-mentored the project and helped guide the progress of the work.

>>>>

\texttt{nls()} is the tool in base R, the primary software software
distribution from the Comprehensive R Archive Network
(\emph{\url{https://cran.r-project.org}}), for estimating nonlinear
statistical models. The function dates to the 1980s and the work related
to Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

@@

\begin{verbatim}
## residual sumsquares =  0  on  11 observations
##     after  1    Jacobian and  1 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## a                     -3            NA         NA         NA           0       26.49  
## b                      5            NA         NA         NA           0       9.615  
## c                    0.5            NA         NA         NA           0       2.466  
## r                    0.7            NA         NA         NA           0      0.1098  
## m                      1            NA         NA         NA           0   1.311e-16
\end{verbatim}

@@

\texttt{nls()}, with the \texttt{numericDeriv()} function, computes the
Jacobian as the ``gradient'' attribute of the residual vector. This is
implemented as a mix of R and C code, but we have created a rather more
compact version entirely in R. See
\emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS}}.

As far as we can understand the logic in \texttt{nls()}, the Jacobian
computation during parameter estimation is carried out entirely within
the called C-language program, and the R code function
\texttt{numericDeriv()}, part of \texttt{./src/library/stats/R/nls.R} in
the R distribution source code. This is used to provide Jacobian
information in the \texttt{nlsModel()} and \texttt{nlsModel.plinear()}
functions, which are \textbf{not} exported for general use.

@@

Note that the selfStart models in the base R
\texttt{./src/library/stats/R/zzModels.R} file and in package
\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
may provide the Jacobian in the ``gradient'' attribute of the
``one-sided'' formula that defines each model, and these Jacobians may
be the analytic forms. The \texttt{nls()} function, after computing the
``right hand side'' or \texttt{rhs} of the residual, then checks to see
if the ``gradient'' attribute is defined, and, if not, uses
\texttt{numericDeriv} to compute a Jacobian into that attribute. This
code is within the \texttt{nlsModel()} or \texttt{nlsModel.plinear()}
functions. The use of analytic Jacobians almost certainly contributes to
the good performance of \texttt{nls()} on selfStart models.

@@

\texttt{nls()}, \texttt{nlsr::nlxb()} and \texttt{minpack.lm::nlsLM()}
all have a \texttt{weights} argument that specifies a vector of weights
the same length as the number of residuals. Each residual is multiplied
by the square root of the corresponding weight. The values returned by
the \texttt{residuals()} function are weighted, and the
\texttt{fitted()} or \texttt{predict()} function are used to compute raw
residuals.

@@

While the implementation of subset at the level of the call to
\texttt{nls()} has a certain attractiveness, it does mean that the
programmer of the solver needs to be aware of the source (and value) of
objects such as the data, residuals and Jacobian. This is overly
complicated. By preference, we would implement subsetting by means of
zero-value weights, with observation counts (and degrees of freedom)
computed via the numbers of non-zero weights. Alternatively, we would
extract a working dataframe from the relevant elements in the original.

@@

\texttt{na.action} is an argument to the \texttt{nls()} function, but it
does not appear obviously in the source code, often being handled behind
the scenes after referencing the option \texttt{na.action}. A useful,
but possibly dated, description is given in:
\emph{\url{https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/}}.

@@

A succinct description of the issue is given in:
\emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}
where the ``Answer'' states

@@

\texttt{nls()} can be called without specifying the \texttt{data}
argument. In this case, it will search in the available environments
(i.e., workspaces) for suitable data objects. We do NOT like this
approach, but it is ``the R way''. R allows users to leave many objects
in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to
actively suppress saving this workspace (\texttt{.RData}) on exit;
otherwise, any such file in the path, when R is launched, will be
loaded. The overwhelming proportion of R users in our acquaintance avoid
saving the workspace because of the danger of lurking data and functions
which may cause unwanted results.

@@

Nonlinear estimation algorithms are almost all iterative and need a set
of starting parameters. \texttt{nls()} offers a special class of
modeling functions called \textbf{selfStart} models. There are a number
of these in base R (\texttt{./src/library/stats/R/zzModels.R}) and
others in R packages such as CRAN package \texttt{nlraa} (Miguez
(\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
now-archived package \texttt{NRAIA}. Unfortunately, some
\textbf{selfStart} codes entangle the calculation of starting values for
parameters with the particulars of the \texttt{nls()} code. Though there
is a \texttt{getInitial()} function, this is not easy to use to simply
compute the initial parameter estimates outside of \texttt{nls()}, in
part because it may call that function. Such circular references are, in
our view, dangerous. Moreover, we believe that it would be helpful to
have selfStart models that allow users to explicitly provide values
other than those suggested for the starting parameters.

@@

By providing starting values that are likely to be reasonable and by
including Jacobian code in the selfStart model code, some of the
deficiencies of the Gauss-Newton algorithm are mitigated. However,
creating such functions is a lot of work, and their documentation (file
\texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated. We
believe that the focus should be placed on getting good initial
parameters, that is \texttt{getInitial()} function, though avoiding the
current calls back to \texttt{nls()}. Interactive tools, such as
``visual fitting'' (Nash and Velleman
(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
considering.

@@

The result object from \texttt{nlsr::nlxb()} is still much smaller than
the one \texttt{nls()} returns. Moreover, \texttt{nlxb} explicitly
returns the sum of squares as well as the residual vector and Jacobian.
The counts of evaluations are also returned. (Note that the singular
values of the Jacobian are actually computed in the \texttt{print} and
\texttt{summary} methods for the result.) As we made our comparisons, We
noted several potential updates to the \texttt{nlsr} documentation as
well as that for \texttt{nls()}.

@@

As already noted, the function \texttt{resid()} (an alias for
\texttt{residuals()}) is WEIGHTED, as are those in the structured object
\texttt{m} returned by \texttt{nls()} or \texttt{minpack.lm::nlsLM},
e.g., \texttt{ans\$m\$resid()}. The function \texttt{nlsModel()}, which
we have had to extract from the base R code and explicitly
\texttt{source()} because it is not exported to the working namespace,
allows us to compute residuals for particular coefficient sets.

@@

As the \texttt{nls()} \textbf{man} page states, when the ``port''
algorithm is used with the \texttt{trace} argument TRUE, the iterations
display the objective function value which is 1/2 the sum of squares (or
deviance). It is likely that the trace display is embedded in the
Fortran of the \texttt{nlminb} routine that is called to execute the
``port'' algorithm, but the factor of 2 discrepancy is nonetheless
unfortunate for users.

@@

Moreover, specifying a model to a solver should, ideally, use the same
syntax across solver tools. Unfortunately, R allows multiple approaches
within different modeling tools, and within \texttt{nls()} itself. This
is highlighted by the case of partially linear models.

@@

\begin{verbatim}
## > with(mm, table(Strip))
## Strip
## S01 S02 S03 
##   4   4   4 
## > nl <- nlevels(mm$Strip)
## > musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, start = list(th = 1), 
## +     algorithm = "plinear")
## > summary(musc.1)
## 
## Formula: Length ~ cbind(1, exp(-Conc/th))
## 
## Parameters:
##       Estimate Std. Error t value Pr(>|t|)    
## th       0.624      0.222    2.81   0.0203 *  
## .lin1   25.684      1.441   17.82  2.5e-08 ***
## .lin2  -26.631      6.147   -4.33   0.0019 ** 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 2.96 on 9 degrees of freedom
## 
## Number of iterations to convergence: 5 
## Achieved convergence tolerance: 4.71e-07
## 
## > b <- coef(musc.1)
## > musc.2 <- nls(Length ~ a[Strip] + b[Strip] * exp(-Conc/th), data = mm, 
## +     start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
## > summary(musc.2)
## 
## Formula: Length ~ a[Strip] + b[Strip] * exp(-Conc/th)
## 
## Parameters:
##    Estimate Std. Error t value Pr(>|t|)    
## a1  22.9277     0.5154   44.49  1.1e-07 ***
## a2  27.8606     0.5150   54.09  4.1e-08 ***
## a3  28.3426     1.0771   26.31  1.5e-06 ***
## b1 -44.1558    12.3900   -3.56    0.016 *  
## b2 -43.8084    12.3148   -3.56    0.016 *  
## b3 -32.8421     2.1553  -15.24  2.2e-05 ***
## th   0.5548     0.0808    6.86    0.001 ** 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.798 on 5 degrees of freedom
## 
## Number of iterations to convergence: 6 
## Achieved convergence tolerance: 1.95e-06
\end{verbatim}

@@

When we started \emph{Improvements to nls()} for the 2021 Google Summer
of Code we wanted:

@@

These goals echo themes in the recent discussion by Vidoni
(\protect\hyperlink{ref-Vidoni21}{2021}) and accompanying commentary.

\hypertarget{rationalization-of-formula-specifications}{%
\subsection{Rationalization of formula
specifications}\label{rationalization-of-formula-specifications}}

We have seen that \texttt{nls()} uses a different formula specification
from the default if the \texttt{plinear} algorithm is used. This is
unfortunate, since the user cannot then simply add
\texttt{algorithm="plinear"} to the standard call, making errors more
likely.

Nonlinear models, apart from those using the ``plinear'' algorithm, need
all the parameters specified. That is, the model formula should be
written down as it would be computed. Thus a linear model (which is
valid input for nonlinear estimation) should be specified
\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} rather than
\texttt{y\ \textasciitilde{}\ x}. Moreover, we want to avoid the
partially linear parameters appearing in the result object as
\texttt{.lin1}, \texttt{.lin2}, etc., where the actual position of
parameters in the model is not explicit. We are unclear why these
parameters begin with a dot, that is, that they have the ``hidden''
indicator.

@@

\hypertarget{rationalization-of-indexed-models}{%
\subsection{Rationalization of indexed
models}\label{rationalization-of-indexed-models}}

@@

\hypertarget{streamlining-code}{%
\subsection{Streamlining code}\label{streamlining-code}}

@@

Such goals align with the aims of \textbf{unit testing} (e.g.,
\emph{\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}},
Wickham (\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
(\protect\hyperlink{ref-HWdevtools21}{2021}) and the conventional R
package testing tools).

\hypertarget{a-testing-package-nlscompare}{%
\subsection{A testing package:
NLSCompare}\label{a-testing-package-nlscompare}}

One of us has developed a working prototype package at
\emph{\url{https://github.com/ArkaB-DS/nlsCompare}} . A primary design
objective of this is to allow the summarisation of multiple tests in a
compact output. The prototype has a vignette to illustrate its use.

@@

In our investigation, we built several resources, which are now part of
the repository \emph{\url{https://github.com/nashjc/RNonlinearLS/}}.
Particular items are:

\begin{itemize}
\item
  A BibTex bibliography for use with all documents in this project, but
  which has wider application to nonlinear least squares projects in
  general
  (\emph{\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib}}).
\item
  \texttt{MachID.R} offers a suggested concise summary function to
  identify a particular computational system used for tests. A
  discussion of how it was built and the resources needed across
  platforms is given in at
  \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}}.
\item
  Nash and Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022})
  is an explanation of the construction of the \texttt{nlspkg} from the
  \texttt{nls()} code in R-base.
\item
  As the 2021 Summer of Code period was ending, one of us (JN) was
  invited to prepare a review of optimization in R. Ideas from the
  present work have been instrumental in the creation of Nash
  (\protect\hyperlink{ref-NashWires22}{2022}).
\end{itemize}

@@

It is possible that \texttt{nls()} will remain more or less as it has
been for the past two decades. Given its importance to R, the focus of
discussion should be the measures needed to secure its continued
operation for legacy purposes and how that may be approached. We welcome
an opportunity to participate in such conversations.

To advance the stability and maintainability of R, we believe the
program objects (R functions) that are created by tools such as
\texttt{nls()} should have minimal cross-linkages and side-effects. The
aspects of \texttt{nls()} that have given us the most trouble:

\begin{itemize}
\item
  The functions that compute the residuals and jacobians often presume
  the data and parameters needed are available in a particular
  environment. As long as the correct environment is used, this provides
  a compact syntax to invoke the calculations. The danger is that the
  wrong data is accessed if the internal search finds a valid name that
  is not the object we want.
\item
  Weights, subsets, and various contextual controls (such as that for
  the \texttt{na.action} option) are similarly taken from the first
  available source, making for a very simple invocation of calculations,
  but the context is hidden from the user. Furthermore, it can be
  difficult for those of us trying to maintain or improve the code to be
  certain we have the context correct.
\item
  The mixing of R, C and Fortran code was necessary for computational
  performance in the past. Lacking decent developer documentation,
  programmers now face a lot of work to fix or improve code. We believe
  in having at least a working reference version of code in a single
  programming language. If necessary, by measuring (``profiling'') the
  code, we can find bottlenecks and substitute for those slower parts of
  the reference code.
\item
  A design that isolates the setup, solution, and post-solution parts of
  for complicated calculations reduces the number of objects that must
  be kept in alignment. On the other hand, it may mean users have to add
  more commands to complete their computations.
\item
  All iterative codes should return the best solution they have found so
  far, even if there are untoward conditions reached, such as a singular
  Jacobian. This modification could be made to \texttt{nls()} in the
  short term, correcting the issue identified above under ``Failure to
  return best result achieved''.
\item
  Given the existence of examples of good practices such as analytic
  derivatives, stabilized solution of Gauss-Newton equations and
  bounds-constrained parameters, base R tools should be moving to
  incorporate them.
\end{itemize}

We anticipate that tools like \texttt{nlsr} will become more attractive
as users discover the increased robustness and general capabilities,
along with the more recent documentation. From a design point of view, a
key difference in approach between \texttt{nls()} and
\texttt{nlsr::nlxb()} is that \texttt{nls()} builds a large
infrastructure, especially the \texttt{nlsModel()} function, from which
the Gauss-Newton iteration can be executed and other statistical
information such as profiles can be computed, while \texttt{nlxb()}
returns quite limited information, and computes what is needed on an
``as and when'' basis. This follows a path that one of us (JN)
established almost 50 years ago with the software that became Nash
(\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
post-solution analysis phases of computations.

@@

Hans Werner Borchers has been helpful in developing the proposal for
this project and in comments on this and related work. Heather Turner
co-mentored the project and helped guide the progress of the work.
Exchanges with Fernando Miguez helped to clarify aspects of selfStart
models and improve package \texttt{nlsr} in that respect.
